The Mysterious Decline in Teen Suicidality in The US during the late 1990s\thanks{Code and data are available at: https://github.com/chrislu2234/suicidality\_report\_1990s}

### 2023-04-22 outputs/paper/paper.tex
Almost everything except for simulated and data tests

---
title: "The Mysterious Decline in Teen Suicidality in The US during the late 1990s"
subtitle: "A Statistical Analysis of the relationship between Legalized Abortion and Teen Suicidality in late 1990s US."
author: 
  - Jin Di Lu
thanks: "Code and data are available at: https://github.com/chrislu2234/suicidality_report_1990s"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This study investigates the correlative relationship between abortion rates and teen suicidality rates in the United States from 1973 to 2000. Using annual time-series data on Abortions and Teen Suicide Rates, I found a significant correlation between abortion rates and teen suicide rates during the studied period, suggesting that the availability of legal abortion might have contributed to the reduction in the number of children born into adverse family conditions and, consequently, alleviated the risk factors for teen suicidality. This research highlights the importance of considering the broader social and psychological implications of abortion policies and calls for further exploration of the causal mechanisms and long-term impacts."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
setwd('C:/Users/chris/OneDrive/Documents/GitHub/suicidality_report_1990s')
library(tidyverse)
library(ggplot2)
library(fable)
library(tsibble)
library(flextable)
```

# Introduction

@@

Several key studies and articles inform my research. Levitt and Donohue's @NBERw25863 seminal work, "The Impact of Legalized Abortion on Crime Over The Last Two Decades," attributes the enactment of Roe v. Wade in 1973 as a major factor contributing to the decline in crime rates across the nation. The authors argue that legalized abortion allowed mothers in emotionally and economically challenging circumstances to make informed choices they couldn't have before, ultimately reducing the number of children born into adverse family conditions to be incentivized towards criminal behavior.

Building on Levitt and Donohue's findings on the impact of legalized abortion on adverse family conditions, in the qualitative study, "Exploring the Family Factors Associated with Suicide Attempts among Adolescents and Young Adults,"@PMC8313455  establishes a connection between the quality of family functioning and relationships and the likelihood of teen suicidality. This connection provides a potential pathway through which legalized abortion might have indirectly contributed to the observed decline in teen suicidality during the 1990s.

Finally, "The Relationship of Family Functioning and Suicidal Ideation among Adolescents: The Mediating Role of Defeat and the Moderating Role of Meaning in Life," @PMC9740712 provide quantitative data supporting the impact of family conditions on teen suicidality. Their research further corroborates the findings of the previously mentioned work and highlights the importance of considering family factors in understanding the decline in teen suicide rates.

@@

This paper uses @citeR, @tidyverse, and @ggplot2 to render, visualize, and clean all the data and graphs used in the paper.

# Data {#sec-data}

@@

To collect data from the desired year ranges (1973-2016), I used three different request forms: Mortality for 1999-2016 with ICD 10 codes, Mortality for 1979-1998 with ICD 9 codes, and Mortality for 1968-1978 with ICD 8 codes. Each request form was submitted to obtain data on suicide deaths among individuals aged 15-19 in the U.S. population, including gender as a variable to identify any differences. The definition of suicide remained consistent across all year ranges as death caused by "self-inflicted harm," despite varying ICD codes. Further information on data retrieval can be found in the readme file.

The Guttmacher Institute's "Pregnancies, Births and Abortions in the United States" report is a periodically updated publication that provides comprehensive data on pregnancy outcomes, including abortions, live births, and miscarriages, in the United States. The report presents vital statistics on abortion rates, ratios, and numbers, as well as data on pregnancy rates and birth rates among different age groups, particularly focusing on women of reproductive age (15-49 years old).

@@

# Models

```{r}
#Read CSVs
setwd('C:/Users/chris/OneDrive/Documents/GitHub/suicidality_report_1990s')
suicides <- read_csv("outputs/clean_data/cleaned_suicides.csv")
abortions <- read_csv("outputs/clean_data/cleaned_abort.csv")
```
Before engaging with the models, it is best to first understand the data. In the exploratory stage of my analysis, I visually represented teen suicide rates between 1973 to 2016. Afterwards, I explored the differences in these rates between males and females, demonstrating to what extent gender affects the statistics in teen suicides.

@@

To engage in proper regression analysis, I also used @fable, @tsibble, and @flextable. Fable offers a wide variety of packages dedicated to forecasting models that use univariate and multivariate time series forecasting models. For this paper, I used a time series linear regression to explore the relationship and influences abortion rates may have had on teen suicide rates. Tsibble was included to help transform my dataframes into Time Series Tibbles that the time series linear regression is able to use. Finally, I included patchwork to improve presentation and overall clarity.

The Time series linear regression model I will be working with is written below:
$$

TSR_{t} = \beta_{0} + \beta_{1} AR_{t}+ \epsilon_{t}

$$
* The Dependent variable being investigated is Teen Suicide Rates, denoted in TSR. 
* Beta0 represents the intercept, which is a point that will be the starting point of the estimate. 
* Beta1 represents the estimate (slope) of the regression, which represents the percent change in TSR for increase in Abortion Rates (denoted as AR)
* Epsilon represents the innovation terms in a Time-Series Linear Model, such as potential errors.

@@

$$

TSR_{t} = \beta_{0} + \beta_{1} AAR_{t}+ \epsilon_{t}

$$
Hence, I produced a second model with an adjusted rate of abortion. Rather than analyzing the relationship between suicide rates and abortion strictly by their numbers in each of those years, I looked to analyze the relationship between teen suicide rates and abortion rates from 17 years prior. I believe this is the best way to represent the impacts abortions have on the quality of life of children born after its legalization in the United States.

@@

In this second model, I was able to demonstrate the relationship between teen suicide rates and abortions, showing how an increase in abortions from 17 years prior decreased teen suicides. Furthermore, the R Squared of this model was a notable ~0.500. I produced 2 additional models to see if this stands with male and female teens as well. These two additional models are:

$$

TSRm_{t} = \beta_{0} + \beta_{1} AAR_{t}+ \epsilon_{t}

$$
$$

TSRf_{t} = \beta_{0} + \beta_{1} AAR_{t}+ \epsilon_{t}

$$
* The Dependent variable being investigated are Teen Suicide Rates for males sand females, denoted in TSRm and TSRf. 
* Beta0 represents the intercept, which is a point that will be the starting point of the estimate. 
* Beta1 represents the estimate (slope) of the regression, which represents the percent change in TSR for increase in Adjusted Abortion Rates (denoted as AR), AAR are 1973 abortion rates placed in 1990 as a way to represent its effects. 
* Epsilon represents the innovation terms in a Time-Series Linear Model, such as potential errors.

@@

Figure 1: Teen Suicide Rates (Crude per 100 000) Total
```{r}
#| echo: false
#| message: false
#| label: fig-1
#| fig-cap: Teen Suicide Rates (Crude per 100 000).
#Filter for Total Suicides from Dataset
suicides_total <- filter(suicides, gender %in% "Total")
#Create Graph
suicides_total %>% 
  ggplot( aes(x=year, y=suicideratetotal)) +
    geom_line( color = "black", size = 1) +
    labs(title= "Suicide Rates of Teens (Crude Rate of Suicides Per 100 000) Between 1973 to 2016", 
         y = "Suicides per 100 000 (Crude)", x = "Year")
```
@fig-1 represents the overall trend of crude teen suicide rates per 100 000 (measured as deaths/population * 100 000) between the years 1973 to 2016. If we look at the plot from before the 2000s, it appears to be nearly unimodal. There is a rise that peaks around the 1990s, and immediately declines in 1995.

@@

```{r}
#| echo: false
#| message: false
#| label: fig-2
#| fig-cap: Teen Suicide Rates (Crude per 100 000) for Males. 
#Filter for Total Suicides from Dataset
suicides_male <- filter(suicides, gender %in% "Male")
#Create Graph
suicides_male %>% 
  ggplot( aes(x=year, y=suicideratetotal)) +
    geom_line( color = "blue", size = 1) +
    labs(title= "Suicide Rates of Teens (Crude Rate of Suicides Per 100 000) Between 1973 to 2006 for Males", 
         y = "Suicides per 100 000 (Crude)", x = "Year")
```
@fig-2 represents crude male teen suicide rates per 100 000 between 1973 to 2016. Similar to @fig-1, this figure also is nearly unimodal before the 2000s. There is a rise that peaks around the 1990s, and immediately declines in 1995. It is important to note here that the rate of suicides amongst male teens is higher than that of the total, indicating (as will be demonstrated by @fig-3 later) that female teen suicide rates are likely much lower.

Figure 3: Teen Suicide Rates (Crude per 100 000) Female
```{r}
#| echo: false
#| message: false
#| label: fig-3
#| fig-cap: Teen Suicide Rates (Crude per 100 000) for Females.
#Filter for Total Suicides from Dataset
suicides_female <- filter(suicides, gender %in% "Female")
#Create Graph
suicides_female %>% 
  ggplot( aes(x=year, y=suicideratetotal)) +
    geom_line( color = "purple", size = 1) +
    labs(title= "Suicide Rates of Teens (Crude Rate of Suicides Per 100 000) for Females",
    y = "Suicides per 100 000 (Crude)", 
    x = "Year")

```
@fig-3 represents crude female teen suicide rates per 100 000 between 1973 to 2016. Contrary to male teen suicides, the unimodality of this figure is not as severe.. Furthermore, the rate is dramatically lower than that of male teens. Nevertheless, there is a rise that peaks around the 1990s, and immediately declines in 1995.

Figure 4: Teen Suicide Rates (Crude per 100 000) Male, Female, Total.
```{r}
#| echo: false
#| message: false
#| label: fig-4
#| fig-cap: Teen Suicide Rates (Crude per 100 000) for Males, Females, and Averaged Total. 
#Create Graph
ggplot() +
  geom_line(data = suicides_female, aes(x = year, y = suicideratetotal), color = "purple", size = 1) +
  geom_line(data = suicides_male, aes(x = year, y = suicideratetotal), color = "blue", size = 1) +
  geom_line(data = suicides_total, aes(x = year, y = suicideratetotal), color = "black", size = 1) +
  labs(title= "Suicide Rates of Teens (Crude Rate of Suicides Per 100 000) for All",
  y = "Suicides per 100 000 (Crude)", 
  x = "Year")
```
@fig-4 represents crude male, female, and all teen suicide rates per 100 000 between 1973 to 2016. As mentioned before, there is a disproportionately high amount of male teen suicides in comparison to females. this averages out in the black bar in the middle, which represents the suicide rates of all teens overall.

Figure 5: Abortion Rates (Number of Abortions per 1 000 Women)
```{r}
#| echo: false
#| message: false
#| label: fig-5
#| fig-cap: Abortion Rates (Number of Abortions per 1 000 Women)
abortions %>% 
  ggplot( aes(x=year, y=abortionratetotal)) +
    geom_line( color = "red", size = 1) +
    ggtitle("Number of Abortions per 1000 Women")
```
@fig-5 represents abortion rates (Number of Abortions per 1000 Women) between 1973 to 2016. Red line follows the increase and decreases of this trend over time. In this figure, we can see the sudden rise in abortion rates after 1973 as a product of the Roe v. Wade Supreme court Decision to Legalize Abortion. After this, we see how this widespread legalization lead to an improvement to accessible abortion resources and a spike in abortion rates. This trend peaks around the 1980s, and follows a slow decline afterwards.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Mysterious Decline in Teen Suicidality in The US during the late 1990s},
  pdfauthor={Jin Di Lu},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{The Mysterious Decline in Teen Suicidality in The US during the
late 1990s\thanks{Code and data are available at:
https://github.com/chrislu2234/suicidality\_report\_1990s}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A Statistical Analysis of the relationship between Legalized
Abortion and Teen Suicidality in late 1990s US.}
\author{Jin Di Lu}
\date{22 April 2023}

\begin{document}
\maketitle
\begin{abstract}
This study investigates the correlative relationship between abortion
rates and teen suicidality rates in the United States from 1973 to 2000.
Using annual time-series data on Abortions and Teen Suicide Rates, I
found a significant correlation between abortion rates and teen suicide
rates during the studied period, suggesting that the availability of
legal abortion might have contributed to the reduction in the number of
children born into adverse family conditions and, consequently,
alleviated the risk factors for teen suicidality. This research
highlights the importance of considering the broader social and
psychological implications of abortion policies and calls for further
exploration of the causal mechanisms and long-term impacts.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Several key studies and articles inform my research. Levitt and
Donohue's (\textbf{NBERw25863?}) seminal work, ``The Impact of Legalized
Abortion on Crime Over The Last Two Decades,'' attributes the enactment
of Roe v. Wade in 1973 as a major factor contributing to the decline in
crime rates across the nation. The authors argue that legalized abortion
allowed mothers in emotionally and economically challenging
circumstances to make informed choices they couldn't have before,
ultimately reducing the number of children born into adverse family
conditions to be incentivized towards criminal behavior.

Building on Levitt and Donohue's findings on the impact of legalized
abortion on adverse family conditions, in the qualitative study,
``Exploring the Family Factors Associated with Suicide Attempts among
Adolescents and Young Adults,''(\textbf{PMC8313455?}) establishes a
connection between the quality of family functioning and relationships
and the likelihood of teen suicidality. This connection provides a
potential pathway through which legalized abortion might have indirectly
contributed to the observed decline in teen suicidality during the
1990s.

Finally, ``The Relationship of Family Functioning and Suicidal Ideation
among Adolescents: The Mediating Role of Defeat and the Moderating Role
of Meaning in Life,'' (\textbf{PMC9740712?}) provide quantitative data
supporting the impact of family conditions on teen suicidality. Their
research further corroborates the findings of the previously mentioned
work and highlights the importance of considering family factors in
understanding the decline in teen suicide rates.

@@

This paper uses R Core Team (2020), Wickham et al. (2019), and Wickham
(2016) to render, visualize, and clean all the data and graphs used in
the paper.

\hypertarget{sec-data}{%
\section{Data}\label{sec-data}}

@@

To collect data from the desired year ranges (1973-2016), I used three
different request forms: Mortality for 1999-2016 with ICD 10 codes,
Mortality for 1979-1998 with ICD 9 codes, and Mortality for 1968-1978
with ICD 8 codes. Each request form was submitted to obtain data on
suicide deaths among individuals aged 15-19 in the U.S. population,
including gender as a variable to identify any differences. The
definition of suicide remained consistent across all year ranges as
death caused by ``self-inflicted harm,'' despite varying ICD codes.
Further information on data retrieval can be found in the readme file.

The Guttmacher Institute's ``Pregnancies, Births and Abortions in the
United States'' report is a periodically updated publication that
provides comprehensive data on pregnancy outcomes, including abortions,
live births, and miscarriages, in the United States. The report presents
vital statistics on abortion rates, ratios, and numbers, as well as data
on pregnancy rates and birth rates among different age groups,
particularly focusing on women of reproductive age (15-49 years old).

@@

\hypertarget{models}{%
\section{Models}\label{models}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Read CSVs}
\FunctionTok{setwd}\NormalTok{(}\StringTok{\textquotesingle{}C:/Users/chris/OneDrive/Documents/GitHub/suicidality\_report\_1990s\textquotesingle{}}\NormalTok{)}
\NormalTok{suicides }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"outputs/clean\_data/cleaned\_suicides.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Rows: 132 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr (2): age group, gender
dbl (4): year, deaths, population, suicideratetotal

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{abortions }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"outputs/clean\_data/cleaned\_abort.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Rows: 44 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
dbl (6): year, abortionstotal, pregnanciestotal, population1544, abortionrat...

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
\end{verbatim}

Before engaging with the models, it is best to first understand the
data. In the exploratory stage of my analysis, I visually represented
teen suicide rates between 1973 to 2016. Afterwards, I explored the
differences in these rates between males and females, demonstrating to
what extent gender affects the statistics in teen suicides.

@@

To engage in proper regression analysis, I also used O'Hara-Wild,
Hyndman, and Wang (2023), Wang, Cook, and Hyndman (2020), and Gohel and
Skintzos (2023). Fable offers a wide variety of packages dedicated to
forecasting models that use univariate and multivariate time series
forecasting models. For this paper, I used a time series linear
regression to explore the relationship and influences abortion rates may
have had on teen suicide rates. Tsibble was included to help transform
my dataframes into Time Series Tibbles that the time series linear
regression is able to use. Finally, I included patchwork to improve
presentation and overall clarity.

The Time series linear regression model I will be working with is
written below: \$\$

TSR\_\{t\} = \beta\emph{\{0\} + \beta}\{1\} AR\_\{t\}+ \epsilon\_\{t\}

\$\$ * The Dependent variable being investigated is Teen Suicide Rates,
denoted in TSR. * Beta0 represents the intercept, which is a point that
will be the starting point of the estimate. * Beta1 represents the
estimate (slope) of the regression, which represents the percent change
in TSR for increase in Abortion Rates (denoted as AR) * Epsilon
represents the innovation terms in a Time-Series Linear Model, such as
potential errors.

@@

\$\$

TSR\_\{t\} = \beta\emph{\{0\} + \beta}\{1\} AAR\_\{t\}+ \epsilon\_\{t\}

\$\$ Hence, I produced a second model with an adjusted rate of abortion.
Rather than analyzing the relationship between suicide rates and
abortion strictly by their numbers in each of those years, I looked to
analyze the relationship between teen suicide rates and abortion rates
from 17 years prior. I believe this is the best way to represent the
impacts abortions have on the quality of life of children born after its
legalization in the United States.

@@

In this second model, I was able to demonstrate the relationship between
teen suicide rates and abortions, showing how an increase in abortions
from 17 years prior decreased teen suicides. Furthermore, the R Squared
of this model was a notable \textasciitilde0.500. I produced 2
additional models to see if this stands with male and female teens as
well. These two additional models are:

\$\$

TSRm\_\{t\} = \beta\emph{\{0\} + \beta}\{1\} AAR\_\{t\}+ \epsilon\_\{t\}

\[
\]

TSRf\_\{t\} = \beta\emph{\{0\} + \beta}\{1\} AAR\_\{t\}+ \epsilon\_\{t\}

\$\$ * The Dependent variable being investigated are Teen Suicide Rates
for males sand females, denoted in TSRm and TSRf. * Beta0 represents the
intercept, which is a point that will be the starting point of the
estimate. * Beta1 represents the estimate (slope) of the regression,
which represents the percent change in TSR for increase in Adjusted
Abortion Rates (denoted as AR), AAR are 1973 abortion rates placed in
1990 as a way to represent its effects. * Epsilon represents the
innovation terms in a Time-Series Linear Model, such as potential
errors.

@@

Figure 1: Teen Suicide Rates (Crude per 100 000) Total

\begin{verbatim}
Warning: Using `size` aesthetic for lines was deprecated in ggplot2 3.4.0.
i Please use `linewidth` instead.
\end{verbatim}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-1-1.pdf}

}

\caption{\label{fig-1}Teen Suicide Rates (Crude per 100 000).}

\end{figure}

Figure~\ref{fig-1} represents the overall trend of crude teen suicide
rates per 100 000 (measured as deaths/population * 100 000) between the
years 1973 to 2016. If we look at the plot from before the 2000s, it
appears to be nearly unimodal. There is a rise that peaks around the
1990s, and immediately declines in 1995.

@@

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-2-1.pdf}

}

\caption{\label{fig-2}Teen Suicide Rates (Crude per 100 000) for Males.}

\end{figure}

Figure~\ref{fig-2} represents crude male teen suicide rates per 100 000
between 1973 to 2016. Similar to Figure~\ref{fig-1}, this figure also is
nearly unimodal before the 2000s. There is a rise that peaks around the
1990s, and immediately declines in 1995. It is important to note here
that the rate of suicides amongst male teens is higher than that of the
total, indicating (as will be demonstrated by Figure~\ref{fig-3} later)
that female teen suicide rates are likely much lower.

Figure 3: Teen Suicide Rates (Crude per 100 000) Female

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-3-1.pdf}

}

\caption{\label{fig-3}Teen Suicide Rates (Crude per 100 000) for
Females.}

\end{figure}

Figure~\ref{fig-3} represents crude female teen suicide rates per 100
000 between 1973 to 2016. Contrary to male teen suicides, the
unimodality of this figure is not as severe.. Furthermore, the rate is
dramatically lower than that of male teens. Nevertheless, there is a
rise that peaks around the 1990s, and immediately declines in 1995.

Figure 4: Teen Suicide Rates (Crude per 100 000) Male, Female, Total.

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-4-1.pdf}

}

\caption{\label{fig-4}Teen Suicide Rates (Crude per 100 000) for Males,
Females, and Averaged Total.}

\end{figure}

Figure~\ref{fig-4} represents crude male, female, and all teen suicide
rates per 100 000 between 1973 to 2016. As mentioned before, there is a
disproportionately high amount of male teen suicides in comparison to
females. this averages out in the black bar in the middle, which
represents the suicide rates of all teens overall.

Figure 5: Abortion Rates (Number of Abortions per 1 000 Women)

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-5-1.pdf}

}

\caption{\label{fig-5}Abortion Rates (Number of Abortions per 1 000
Women)}

\end{figure}

Figure~\ref{fig-5} represents abortion rates (Number of Abortions per
1000 Women) between 1973 to 2016. Red line follows the increase and
decreases of this trend over time. In this figure, we can see the sudden
rise in abortion rates after 1973 as a product of the Roe v. Wade
Supreme court Decision to Legalize Abortion. After this, we see how this
widespread legalization lead to an improvement to accessible abortion
resources and a spike in abortion rates. This trend peaks around the
1980s, and follows a slow decline afterwards.

Graph 6: Graph Teen Suicide Rates vs Abortion Rates

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-6-1.pdf}

}

\caption{\label{fig-6}Teen Suicide Rates vs.~Abortion Rates}

\end{figure}

Figure~\ref{fig-6} represents an attempt to plot teen suicide rates
against abortion rates. However, this plot provides very little
information on the topic, as the rates are not proportional to one and
another. But if we look closely, the dramatic increase in abortion rates
in 1973 can be argued to have manifested in the sharp declinee in teeen
suicides in 1995.

\hypertarget{results}{%
\section{Results}\label{results}}

Model 1: Teen Suicide Rates vs.~Abortion Rates

\begin{verbatim}
Series: suicideratetotal 
Model: TSLM

Residuals:
    Min      1Q  Median      3Q     Max 
-1.8699 -1.1347 -0.2501  1.4252  1.9051

Coefficients:
                  Estimate Std. Error t value Pr(>|t|)   
(Intercept)         5.4416     1.8209   2.988  0.00605 **
abortionratetotal   0.1499     0.0718   2.088  0.04672 * 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 1.273 on 26 degrees of freedom
Multiple R-squared: 0.1436, Adjusted R-squared: 0.1107
F-statistic: 4.361 on 1 and 26 DF, p-value: 0.04672
\end{verbatim}

\hypertarget{tbl-1}{}
\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}
